---
title: "R Fundamentals - Part 1 of 4: Introduction"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r}
library(here)
source(here("source","get_lib.R"))
```

# Writing and Running Code

```{r}
# ctrl + alt + i  : code chunk

# Inside code chunks, text lines need to be started with a # (hashtag)

# You can use R like a giant calculator! 
a <- 2 + 3
```

All variables created show up in the Global Environment (top right panel).

```{r}
# print a list of all variables in your environment
ls()

# remove a single variable
rm(a)

```

```{r}
# variable storing the population of Mexico
pop <- 126

# rename to be more informative
pop_millions <- pop

# print it to confirm that it holds the value of pop
pop_millions

# remove the old variable
rm(pop)

# print out a list of current variables
ls()

# remove all variables in the global environment
rm(list = ls())
```

💡 **Tip:** `ls()` and `rm()` are both **functions**, which are commands that generate some output. We will learn several functions across the workshop. The best way to learn and remember functions in R is to Google them or ask ChatGPT!

### 🥊 Challenge 3: Creating Variables from Other Variables

Pick a country. Look up its population and create a variable which stores the population. Now look up the GDP (Gross Domestic Product) and create another variable which stores the GDP.

Can you use these two variables to create a third variable which stores the GDP per capita (GDP divided by population)?

```{r}



```

## Creating Vectors

Most of the time, we want to store multiple values. For example, I might want to record the population in Mexico from 2018 to 2022. **Vectors** can store multiple values as long as they're the same type (all numeric, all characters etc).

To create a vector, you use the `c()` function, which stands for concatenate (meaning combine).

```{r}
# create a vector with the approx population of Mexico for 5 years  
c(123, 124, 125, 126, 127)

# as chance would have it, Mexico's population has increased by about 1 million each year from 2018 to 2022
```

Vectors are assigned to variables using the same assignment operator (\<-) that we used for single value variables.

```{r}
# save the population counts to a vector
pop_mexico <- c(123, 124, 125, 126, 127)
pop_mexico
```

If we do something to the values in a vector, R applies that operation to each item in the vector.

```{r}
# multiple the values in pop_mexico by 1 million
pop_mexico*1000000

# you forgot to count 2 million people - add them to our counts
pop_mexico + 2
```

## 🎬 **Demo**: Next Time...

Most of the time, you won't be working with single value variables or vectors. Instead, you'll be working with data that contains multiple variables. In Part 2, we will learn how to read in data to R and create **data frames** (a way to store data with multiple rows and columns).

🔔 **Tip:** Run the two chunks of code below. What do you think we are doing?

```{r}
# read in data
gap <- read.csv(file = "../data/gapminder.csv")
head(gap)

```

```{r}
# subset data
gap[gap$country == "Mexico", c("country", "year", "pop")] 


```

## Key Points

-   In RStudio you can write code in an **R Notebook** or an **R Script** - we are using a notebook

-   Use the syntax **variable \<- value** to create a variable and assign it a value

-   Vectors store multiple values as long as they're the same type

-   We used the following functions: `ls()`, `rm()`, and `c()`. To see how to use them use **?** followed by the function name (or Google them)
